# Univariate portfolio sorts

In this chapter, we dive into portfolio sorts, one of the most widely used statistical methodologies in empirical asset pricing. The key application of portfolio sorts is to examine whether one or more variables can predict future excess returns. In general, the idea is to sort individual stocks into portfolios, where the stocks within each portfolio are similar with respect to a sorting variable, such as firm size. The different portfolios then represent well-diversified investments that differ in the level of the sorting variable. You can then attribute the differences in the return distribution to the impact of the sorting variable. 
We start by introducing univariate portfolio sorts (which sort based on only one characteristic). Later, we tackle bivariate sorting. 

A univariate portfolio sort considers only one sorting variable $x_{t-1,i}$. Here, $i$ denotes the stock and $t-1$ indicates that the characteristic is observable by investors at time $t$.  
The objective is to assess the cross-sectional relation between $x_{t-1,i}$ and, typically, stock excess returns $r_{t,i}$ at time $t$ as the outcome variable. To illustrate how portfolio sorts work, we use estimates for market betas from the previous chapter as our sorting variable.

## Data preparation

The current chapter relies on the following set of packages.

```{r}
library(tidyverse)
library(RSQLite)
library(lubridate)
library(sandwich)
library(lmtest)
library(scales)
```

We start with loading the required data from our `SQLite`-database. In particular, we use the monthly CRSP database as our asset universe. Once we form our portfolios, we use the Fama-French factor returns to compute the risk-adjusted performance (i.e., alpha). `beta` is the tibble with equity betas computed in the previous chapter. 

```{r}
tidy_finance <- dbConnect(SQLite(), "data/tidy_finance.sqlite", extended_types = TRUE)

crsp_monthly <- tbl(tidy_finance, "crsp_monthly") %>% 
  collect() 

factors_ff_monthly <- tbl(tidy_finance, "factors_ff_monthly") %>% 
  collect()

beta <- tbl(tidy_finance, "beta") %>% 
  collect()
```

We keep only relevant data from the CRSP sample.

```{r}
crsp_monthly <- crsp_monthly %>% 
  left_join(factors_ff_monthly, by = "month") %>% 
  select(permno, month, ret_excess, mkt_excess, mktcap_lag)
crsp_monthly
```

## Sorting by market beta

Next, we merge our sorting variable with the return data. We use the one-month *lagged* betas as a sorting variable to ensure that the sorts rely only on information available when we create the portfolios. To lag stock beta by one month, we add one month to the current date and join the resulting information with our return data. This procedure ensures that month $t$ information is available in month $t+1$. You may be tempted to simply use a call such as `crsp_monthly %>% group_by(permno) %>% mutate(beta_lag = lag(beta)))` instead. This procedure, however, does not work if there are non-explicit missing values in the time series.

```{r}
beta_lag <- beta %>%
  mutate(month = month %m+% months(1)) %>% 
  select(permno, month, beta_lag = beta_daily) %>%
  drop_na()

data_beta <- crsp_monthly %>%
  inner_join(beta_lag, by = c("permno", "month"))
```

The first step of the portfolio analysis is to calculate periodic breakpoints that you can use to group the stocks into portfolios. For simplicity, we start with the median as the single breakpoint. We then compute the value-weighted returns for each of the two resulting portfolios which means that the lagged market capitalizations determines the weight in `weighted.mean()`.  

```{r}
beta_portfolios <- data_beta %>%
  group_by(month) %>%
  mutate(breakpoint = median(beta_lag),
         portfolio = case_when(beta_lag <= breakpoint ~ "low",
                               beta_lag > breakpoint ~ "high")) %>%
  group_by(month, portfolio) %>%
  summarize(ret = weighted.mean(ret_excess, mktcap_lag), .groups = "drop")
```


## Performance evaluation (I)

The following figure shows the monthly excess returns of the two portfolios.

```{r}
beta_portfolios %>% 
  ggplot(aes(x = month, y = ret, fill = portfolio)) +
  geom_col() +
  facet_wrap(~portfolio, ncol = 1) +
  scale_y_continuous(labels = percent) + 
  labs(x = NULL, y = NULL, 
       title = "Monthly beta portfolio excess returns using median as breakpoint") +
  theme_bw() +
  theme(legend.position = "none")
```

We can construct a long-short strategy based on the two portfolios: buy the high-beta portfolio and, at the same time, short the low-beta portfolio. Thereby, the overall position in the market is net zero, i.e., you do not need invest money to realize this strategy in the absence of frictions.

```{r}
beta_longshort <- beta_portfolios %>%
  pivot_wider(month, names_from = portfolio, values_from = ret) %>%
  mutate(long_short = high - low) %>% 
  left_join(factors_ff_monthly, by = "month") 
```

We compute the average return and the corresponding standard error to test whether the long-short portfolio yields on average positive or negative excess returns. In the asset pricing literature, one typically uses [@Newey1987] $t$-statistics to test the null hypothesis that average portfolio excess returns are equal to zero. To implement this test, we compute the average return via `lm()` and then employ the `coeftest` function.

```{r}
model_fit <- lm(long_short ~ 1, data = beta_longshort)
coeftest(model_fit, vcov = NeweyWest, lag = 6)
```

The results indicate that we cannot reject the null hypothesis of average returns being equal to zero. Our portfolio strategy using the median as a breakpoint hence does not yield any abnormal returns. Is this finding surprising if you reconsider the CAPM? It certainly is. The CAPM yields that the high beta stocks should yield higher expected returns. Our portfolio sort implicitly mimics an investment strategy that finances high beta stocks by shorting low beta stocks. Therefore, one should expect that the average excess returns yield a return that is above the risk-free rate.


## Functional programming for portfolio sorts

Now we take portfolio sorts to the next level. We want to be able to sort stocks into an arbitrary number of portfolios. For this case, functional programming is very handy: we employ the [curly-curly](https://www.tidyverse.org/blog/2019/06/rlang-0-4-0/#a-simpler-interpolation-pattern-with-)-operator to give us flexibility concerning which variable to use for the sorting, denoted by `var`. We use `quantile()` to compute breakpoints for `n_portfolios`. Then, we assign portfolios to stocks using the `findInterval()` function. The output of the following function is a new column that contains the number of the portfolio to which a stock belongs.

```{r}
assign_portfolio <- function(data, var, n_portfolios) {
  breakpoints <- data %>%
    summarize(breakpoint = quantile({{ var }}, probs = seq(0, 1, length.out = n_portfolios + 1),
                                    na.rm = TRUE)) %>%
    pull(breakpoint) %>% 
    as.numeric()
  
  data %>%
    mutate(portfolio = findInterval({{ var }}, breakpoints, all.inside = TRUE)) %>% 
    pull(portfolio)
}
```

We can use the above function to sort stocks into ten portfolios each month using lagged betas and compute value-weighted returns for each portfolio. Note that we transform the portfolio column to a factor variable because it provides more convenience for the figure construction below.

```{r}
beta_portfolios <- data_beta %>%
  group_by(month) %>%
  mutate(portfolio = assign_portfolio(data = cur_data(), 
                                      var = beta_lag, 
                                      n_portfolios = 10),
         portfolio = as.factor(portfolio)) %>%
  group_by(portfolio, month) %>% 
  summarize(ret = weighted.mean(ret_excess, mktcap_lag), .groups = "drop")
```


## Performance evaluation (II)

In the next step, we compute summary statistics for each beta portfolio. Namely, we compute CAPM-adjusted alphas, the beta of each beta portfolio, and average returns. 

```{r}
beta_portfolios_summary <- beta_portfolios %>%
  left_join(factors_ff_monthly, by = "month") %>%
  group_by(portfolio) %>%
  summarise(alpha = as.numeric(lm(ret ~ 1 + mkt_excess)$coefficients[1]),
            beta = as.numeric(lm(ret ~ 1 + mkt_excess)$coefficients[2]),
            ret = mean(ret))
```

The figure below illustrates the CAPM alphas of beta-sorted portfolios. It shows that low beta portfolios tend to exhibit positive alphas, while high beta portfolios exhibit negative alphas.

```{r}
beta_portfolios_summary %>% 
  ggplot(aes(x = portfolio, y = alpha, fill = portfolio)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(title = "Alphas of beta-sorted portfolios",
       x = "Portfolio",
       y = "CAPM alpha",
       fill = "Portfolio") +
  theme_bw() +
  scale_y_continuous(labels = percent)
```

These results suggest a negative relation between beta and future stock returns, which contradicts the predictions of the CAPM. According to the CAPM, returns should increase with beta across the portfolios and risk-adjusted returns should be statistically indistinguishable from zero.


## The security market line and beta portfolios

The CAPM predicts that our portfolios should lie on the security market line (SML). The slope of the SML is equal to the market risk premium and reflects the risk-return trade-off at any given time.

```{r}
sml_capm <- lm(ret ~ 1 + beta, data = beta_portfolios_summary)$coefficients 

beta_portfolios_summary %>% 
  ggplot(aes(x = beta, y = ret, color = portfolio)) +
  geom_point() +
  geom_abline(intercept = 0, slope = mean(factors_ff_monthly$mkt_excess)) +
  geom_abline(intercept = sml_capm[1], slope = sml_capm[2], color = "green") +
  scale_y_continuous(labels = percent, limit = c(0, mean(factors_ff_monthly$mkt_excess)*2)) +
  scale_x_continuous(limits = c(0,2)) +
  theme_bw() +
  labs(x = "Beta", y = "Excess return", color = "Portfolio",
       title = "Average portfolio excess returns and average beta estimates")
```

To provide more evidence against the CAPM predictions, we again form a long-short strategy that buys the high-beta portfolio and shorts the low-beta portfolio. 

```{r}
beta_longshort <- beta_portfolios %>%
  ungroup() %>%
  mutate(portfolio = case_when(portfolio == max(as.numeric(portfolio)) ~ "high",
                               portfolio == min(as.numeric(portfolio)) ~ "low")) %>% 
  filter(portfolio %in% c("low", "high")) %>%
  pivot_wider(month, names_from = portfolio, values_from = ret) %>%
  mutate(long_short = high - low) %>% 
  left_join(factors_ff_monthly, by = "month") 
```

Again, the resulting long-short strategy does not exhibit statistically significant returns. 

```{r}
coeftest(lm(long_short ~ 1, data = beta_longshort), vcov = NeweyWest)
```

However, the long-short portfolio yields a statistically significant negative CAPM-adjusted alpha, although, controlling for the effect of beta, the average excess stock returns should be zero according to the CAPM. The results thus provide no evidence in support of the CAPM. The negative value has been documented as the so-called betting against beta factor ([@Frazzini2014]). Betting against beta corresponds to a strategy that shorts high beta stocks and takes a (levered) long position in low beta stocks. If borrowing constraints prevent investors from taking positions on the SML they are instead incentivized to buy high beta stocks, which leads to a relatively higher price (and therefore lower expected returns than implied by the CAPM) for such high beta stocks. As a result, the betting-against-beta strategy earns from providing liquidity to capital constraint investors with lower risk aversion. 

```{r}
coeftest(lm(long_short ~ 1 + mkt_excess, data = beta_longshort), vcov = NeweyWest)
```

The plot below shows the annual returns of the extreme beta portfolios we are mainly interested in. The figure illustrates no consistent striking patterns over the last years - each portfolio exhibits periods with positive and negative annual returns. 

```{r}
beta_longshort %>% 
  group_by(year = year(month)) %>% 
  summarize(low = prod(1 + low),
            high = prod(1 + high),
            long_short = prod(1 + long_short)) %>% 
  pivot_longer(cols = -year) %>% 
  ggplot(aes(x = year, y = 1 - value, fill = name)) +
  geom_col(position = "dodge") + 
  facet_wrap(~name, ncol = 1) +
  theme_bw() + theme(legend.position = "none") + 
  scale_y_continuous(labels = percent) + 
  labs(title = "Annual returns of beta portfolios",
       x = NULL, y = NULL)
```

Overall, this chapter shows how functional programming can be leveraged to form an arbitrary number of portfolios using any sorting variable and how to evaluate the performance of the resulting portfolios. In the next chapter, we dive deeper into the many degrees of freedom that arise in the context of portfolio analysis. 


## Exercises

1. Take the two long-short beta strategies based on different numbers of portfolios and compare the returns. Is there a significant difference in returns? How do the Sharpe ratios compare between the strategies? Find one additional portfolio evaluation statistic and compute it.
1. We plotted the alphas of the ten beta portfolios above. Write a function that tests these estimates for significance. Which portfolios have significant alphas?
1. The analysis here is based on betas from daily returns. However, we also computed betas from monthly returns. Re-run the analysis and point out differences in the results.
1. Given the results in this chapter, can you define a long-short strategy that yields positive abnormal returns (i.e., alphas)? Plot the cumulative excess return of your strategy and the market excess return for comparison.
