<!-- Replace correlation test and t test by regressions and interpret them (+ drop lmtest package?) -->
<!-- Exercise: create typical table for portfolio sorts for each portfolio used in FF construction -->
<!-- exercise: compute FF alphas for beta portfolios from chapter 32 -->

# Replicating Fama & French factors

The Fama and French three-factor model (see [@Fama1993]) is a cornerstone of asset pricing. On top of the market factor represented by the traditional CAPM beta, the model includes the factors size and value. We did introduce both factors in the previous chapter, and their definition remains the same. Size is the SMB factor (small-minus-big) that is long small firms and short large firms. The value factor is HML (high-minus-low) and is long in high book-to-market firms and short the low book-to-market counterparts. While we demonstrated the main idea of portfolio sorts, we also want to show how to replicate these significant factors. However, we do not aim at a perfect replication but want to show the main ideas.

The current chapter relies on this set of packages. 
```{r}
library(tidyverse)
library(RSQLite)
library(lubridate)
library(sandwich)
library(lmtest)
```

## Databases

We use the same data sources CRSP and compustat, as we need exactly the same variables to compute the size and value factors in the way Fama and French do it. Hence, there is nothing new below.

```{r}
tidy_finance <- dbConnect(SQLite(), "data/tidy_finance.sqlite", extended_types = TRUE)

crsp_monthly <- tbl(tidy_finance, "crsp_monthly") %>% 
  collect()

factors_ff_monthly <- tbl(tidy_finance, "factors_ff_monthly") %>% 
  collect()

compustat <- tbl(tidy_finance, "compustat") %>% 
  collect()

data_ff <- crsp_monthly %>% 
  left_join(factors_ff_monthly, by = "month") %>% 
  select(permno, gvkey, month, ret_excess, mkt_excess, mktcap, mktcap_lag, exchange) %>% 
  drop_na()

be <- compustat %>%
  select(gvkey, datadate, be) %>% 
  drop_na()
```

## Data preparation

Yet when we start merging our data set for computing the premiums, there are a few differences to the previous chapter. First, Fama and French form their portfolios in June of year $t$, whereby the returns of July are the first monthly return for the respective portfolio. For firm size, they consequently use the market capitalization recorded for June. It is then held constant until June of year $t+1$.

Second, Fama and French also have a different protocol for computing the book-to-market ratio. They use market equity as of the end of year $t - 1$ and the book equity reported in year $t-1$, i.e., the `datadate` is within the last year. Hence, the book-to-market ratio can be based on accounting information that is up to 18 months old. Market equity also does not necessarily reflect the same time point as book equity.

To implement all these time lags, we again employ the temporary `sorting_date`-column. Notice that when we combine the information, we want to have a single observation per year and stock since we are only interested in computing the breakpoints held constant for the entire year. We ensure this by a call of `distinct()` at the end of the chunk below.

```{r}
ff_me <- data_ff %>%
  filter(month(month) == 6) %>%
  mutate(sorting_date = month %m+% months(1)) %>%
  select(permno, sorting_date, ff_me = mktcap)

ff_me_dec <- data_ff %>%
  filter(month(month) == 12) %>%
  mutate(sorting_date = ymd(paste0(year(month) + 1, "0701)"))) %>%
  select(permno, gvkey, sorting_date, bm_me = mktcap)

ff_bm <- be %>%
  mutate(sorting_date = ymd(paste0(year(datadate) + 1, "0701"))) %>%
  select(gvkey, sorting_date, bm_be = be) %>%
  drop_na() %>%
  inner_join(ff_me_dec, by = c("gvkey", "sorting_date")) %>%
  mutate(ff_bm = bm_be/bm_me) %>%
  select(permno, sorting_date, ff_bm)

ff_variables <- ff_me %>% 
  inner_join(ff_bm, by = c("permno", "sorting_date")) %>%
  drop_na() %>%
  distinct(permno, sorting_date, .keep_all = TRUE)
```

## Portfolio sorts

Next, we construct our portfolios with an adjusted `assign_portfolio()` function. Fama and French rely on NYSE-specific breakpoints, they form two portfolios in the size dimension at the median and three portfolios in the dimension of book-to-market at the 30%- and 70%-percentiles, and they use independent sorts. The sorts for book-to-market require an adjustment to the previous function, because the `seq()` we would produce does not produce the right breakpoints. Instead of `n_portfolios` we now specify `percentiles`, which take the breakpoint-sequence as an object specified in the function's call. Specifically, we give `percentiles = c(0, 0.3, 0.7, 1)` to the function. Additionally, we perform an `inner_join()`  with our return data to ensure that we only use traded stocks when computing the breakpoints as a first step. 

```{r}
assign_portfolio <- function(data, var, percentiles) {
  breakpoints <- data %>%
    filter(exchange == "NYSE") %>%
    summarize(breakpoint = quantile(
      {{ var }},
      probs = {{ percentiles }},
      na.rm = TRUE
    )) %>%
    pull(breakpoint) %>%
    as.numeric()

  data %>%
    mutate(portfolio = findInterval({{ var }}, breakpoints, all.inside = TRUE)) %>%
    pull(portfolio)
}

ff_portfolios <- ff_variables %>%
  inner_join(data_ff, by = c("permno" = "permno", "sorting_date" = "month")) %>%
  group_by(sorting_date) %>%
  mutate(
    portfolio_me = assign_portfolio(
      data = cur_data(),
      var = ff_me,
      percentiles = c(0, 0.5, 1)
    ),
    portfolio_bm = assign_portfolio(
      data = cur_data(),
      var = ff_bm,
      percentiles = c(0, 0.3, 0.7, 1)
    )) %>%
  select(permno, sorting_date, portfolio_me, portfolio_bm)
```

Next, we merge the portfolios to the return data of the rest of the year. To implement this step, we create a new column `sorting_date` in our return data by setting the date to sort on to July of $t-1$ if the month is June (of year $t$) or earlier or to July of year $t$ if the month is July or later.

```{r}
ff_portfolios <- data_ff %>%
  mutate(sorting_date = case_when(month(month) <= 6 ~ ymd(paste0(year(month) - 1, "0701")),
                                  month(month) >= 7 ~ ymd(paste0(year(month), "0701")))) %>%
  inner_join(ff_portfolios, by = c("permno", "sorting_date"))
```


## Fama and French factor returns

Equipped with the return data and the assigned portfolios, we can now compute the value-weighted average return for each of the six portfolios. Then, we form the Fama and French factors. For the size factor (i.e., *SMB*), we go long in the three small portfolios and short the three large portfolios by taking an average across either group. For the value factor (i.e., *HML*), we go long in the two high book-to-market portfolios and short the two low book-to-market portfolios, again weighting them equally.

```{r}
ff_factors <- ff_portfolios %>%
  mutate(portfolio = paste0(portfolio_me, portfolio_bm)) %>%
  group_by(portfolio, month) %>%
  summarise(ret = weighted.mean(ret_excess, mktcap_lag), .groups = "drop",
            portfolio_me = unique(portfolio_me),
            portfolio_bm = unique(portfolio_bm)) %>%
  group_by(month) %>%
  summarise(ff_SMB = mean(ret[portfolio_me == 1]) - mean(ret[portfolio_me == 2]),
            ff_HML = mean(ret[portfolio_bm == 3]) - mean(ret[portfolio_bm == 1]))
```


## Replication evaluation

In the previous section, we replicated the size and value premiums following the procedure outlined by Fama and French. However, we did not follow their procedure strictly. The final question is then; how close did we get? We answer this question by looking at the two time-series estimates in a correlation analysis using the function `cor.test()` and a test of means using the function `t.test()`.


```{r}
test <- factors_ff_monthly %>% 
  inner_join(ff_factors, by = "month") %>%
  mutate(ff_SMB = round(ff_SMB, 4),
         ff_HML = round(ff_HML, 4))

cor.test(test$ff_SMB, test$smb) 
cor.test(test$ff_HML, test$hml)
```


```{r}
t.test(test$ff_SMB, test$smb) 
t.test(test$ff_HML, test$hml)
```
We can see that we reject the null hypothesis for both premiums in the correlation tests, i.e., the hypothesis that the time series are uncorrelated is rejected. Moreover, we cannot reject the hypothesis that the means of the Fama and French factors and our replication results are different. Hence, we can conclude that we did a relatively good job in replicating their premiums.
