# Fama-MacBeth Regressions

The regression approach of @Fama1973 is widely used in empirical asset pricing studies. Researchers use the two-stage regression approach to estimate risk premiums in various markets, but predominately in the stock market. In the end, Fama-MacBeth regressions are used in much the same way as portfolio sorts introduced before.
We present a simple implementation of @Fama1973 to introduce the concept of their regressions.

Frequently, the application of @Fama1973 regressions starts with the estimation of $\beta^{F_f}_{i}$ for some risk factor $F_f$ from a group of $F$ risk factors $F_1, \cdots, F_F$ for each asset $i \in [1; N]$. 
As an example, we consider the three risk factors of @Fama1993 below and estimate the betas from $N$ time-series regressions
$$r_{i,t} = \alpha_i + \beta^M_i r^M_t + \beta^{SMB}_i r^{SMB}_t + \beta^{HML}_i r^{HML}_t + \epsilon_{i,t}\text{, for each } i.$$ We use excess returns as our dependent variable, but the approach could also incorporate raw returns which leads to a change in the interpretation of intercepts (and their standard errors have to be adjusted following @Shanken1992). 

You want to think about the $\beta^{F_f}_{i}$ as an asset characteristic. In fact, we could also use asset characteristics like a firm's size or book-to-market ratio directly in the cross-sectional regression (i.e., the next step) and skip the time-series regressions for such variables. Moreover, we could also consider rolling-window estimation for betas or more evolved forecasts of betas if exposures might vary over time. For example, @Fama1973 use 60-month windows for the estimation of exposures. Other authors use 24-month or 36-month windows for monthly returns or 240 trading days for data on daily frequency. We would add a subscript $t$ to the coefficients in these cases, but we leave this as an exercise.

The second step in the Fama-MacBeth procedure uses the exposures or characteristics in $T$ cross-sectional regressions, i.e.,
$$r_{i,t} = \alpha_i + \lambda^{M}_t \beta^M_i  + \lambda^{SMB}_t \beta^{SMB}_i + \lambda^{HML}_t \beta^{HML}_i + \epsilon_{i,t}\text{, for each t},$$ where we are interested in the compensation $\lambda^{F_f}_t$ for the exposure to each risk factor in each time point, i.e., the risk premium.
The time-series average of the $\lambda^{F_f}_t$ can then be interpreted as the risk premium for the specific risk factor $F_f$.

What about the standard errors of the risk premiums? You might worry that the errors of the $\beta^{F_f}_{i}$ estimated at the beginning impact the risk premium's standard errors. It turns out that you can simply run a T-test on the time-series of $\lambda^{F_f}_t$ to find the standard error of the mean. If you worry about serial correlation in the risk premiums, you can also use the corrected standard errors of @Newey1987. Overall, this is really convenient.

The final remark before moving to the implementation concerns the coefficient estimates themselves. Measurement error in $\beta^{F_f}_{i}$ affects the risk premium estimates, i.e., they lead to biased estimates. The literature provides adjustments for this (see, e.g., @shanken1992, @kim1995, @chen2015, among others), but also shows that the bias goes to zero as $T \to \infty$. @gagliardini2016 provide an indepth discussion covering also the case of time-varying betas and estimation on individual stocks.

A side note, if you plan on using Fama-MacBeth regressions with individual stocks; @hou2020 advocate using weighed-least squares to estimate the coefficients such that they are not biased towards small firms. Without this adjustment, the numerous small firms would drive the coefficient estimates.


The current chapter relies on this set of packages. 
```{r}
library(tidyverse)
library(RSQLite)
library(lubridate)
```

## Data preparation

We can illustrate @Fama1973-regressions with publicly available data from Prof. French's website. We collect the industry portfolio returns, and the three Fama-French factors from our `SQLite`-database introduced in our chapter on *"Accessing & managing financial data"*.

```{r}
tidy_finance <- dbConnect(SQLite(), "data/tidy_finance.sqlite",
  extended_types = TRUE
)

industry_returns <- tbl(tidy_finance, "industries_ff_monthly") %>% 
  collect() 

factors_ff_monthly <- tbl(tidy_finance, "factors_ff_monthly") %>%
  collect()
```

We compute excess returns for the ten test assets by subtracting the risk-free rate with this data at hand. 
We also merge the three Fama-French risk factors into the data to prepare for the next step. We finally rename the market excess return to be consistent for aesthetic purposes.

```{r}
data_famamacbeth <- industry_returns %>%
  left_join(factors_ff_monthly, by = "month") %>%
  mutate(across(NoDur:Other, ~ . - rf)) %>%
  select(-rf)
```

If you worry about the number of test assets being too small, we agree with your concern. Yet this chapter is about introducing Fama-MacBeth regressions as a tool and not about interpreting meaningful results.

## Time-series regression: Factor exposure

The first step in many applications of Fama-MacBeth regressions is the estimation of factor betas. In our case, we have to estimate three $\beta_i$ for the three risk factors of @Fama1993. We do this by writing a function for the estimation (similar to the function in the chapter on *"Beta estimation"*) and then estimating the slope coefficients in one step. The function itself is sufficiently flexible to handle more or fewer risk factors and test assets. 

```{r}
estimate_factor_exposure <- function(asset_return, risk_factors) {
  fit <- lm(as.numeric(asset_return) ~ as.matrix(risk_factors))
  
  table <- as_tibble(as.list(fit$coefficients[2:(ncol(risk_factors) + 1)]))
  colnames(table) <- paste0("beta_", colnames(risk_factors))
  return(table)
}

risk_exposures <- data_famamacbeth %>%
  pivot_longer(cols = NoDur:Other, 
               names_to = "asset", 
               values_to = "excess_return") %>%
  group_by(asset) %>%
  summarize(exposures = factor_exposure(asset_return = cur_data() %>% pull(excess_return), 
                                        risk_factors = cur_data() %>% select(mkt, smb, hml))) %>%
  unpack(exposures)

risk_exposures
```

We now have three factor exposures for each of the ten test assets and are ready to turn our attention to the cross-sectional regression. At this point, you might want to reflect on the exposure estimates and analyze them in more detail. We skip this step to keep it simple.


## Cross-sectional regression: Risk premiums

To smoothly run the cross-sectional regression, we build another function as the one for the time-series estimation of risk exposures. This time, we regress the returns of the test assets at a particular point of time on the factor exposures of each asset. By doing so, we get an estimate of the risk premiums $\lambda^{F_f}_t$ for each point in time.

```{r}
risk_premiums <- function(asset_returns, exposures) {
  asset_returns <- asset_returns %>%
    select(exposures$asset) %>%
    as.numeric()

  exposures <- exposures %>%
    select(-asset) %>%
    as.matrix() 
  
  fit <- lm(asset_returns ~ exposures)
  
  table <- as_tibble(as.list(fit$coefficients)) 
  colnames(table) <- c("alpha", gsub("beta_", "lambda_", colnames(exposures)))
  table
}

data_risk_premiums <- data_famamacbeth %>%
  group_by(month) %>%
  summarize(risk_premiums = risk_premiums(asset_returns = cur_data() %>% select(risk_exposures$asset), 
                exposures = risk_exposures)) %>%
  unpack(risk_premiums)
```

Now that we have the risk premiums' estimates for each period, we can average across the time-series dimension to get the expected risk premium for each risk factor.

```{r}
data_risk_premiums %>%
  select(starts_with("lambda_")) %>%
  summarize(across(everything(), ~ c(t.test(.x)$estimate, 
                                     t.test(.x)$statistic))) %>%
  bind_cols(tibble("Statistc" = c("Mean", "T-value")), .)
```

In this illustration, we can conclude that there is no significant risk premium for the three @Fama1993 factors. However, we would not consider this a valid or exciting answer given the limitations of the setup used here.

## Exercises

1. Download a larger sample of test assets from Kenneth French's homepage and reevaluate the risk premiums.
1. Implement a rolling-window regression for the time-series estimation of the factor exposure. Skip one month after each rolling period before including the exposures in the cross-sectional regression to avoid a look-ahead bias. Then, adapt the cross-sectional regression and compute the average risk premiums.