# Fama-MacBeth Regressions

The regression approach of @Fama1973 is widely used in empirical asset pricing studies. 
Researchers use the two-stage regression approach to estimate risk premiums in various markets, but predominately in the stock market. 
In the end, Fama-MacBeth regressions are used in much the same way as portfolio sorts introduced in previous chapters.
Essentially, Fama-MacBeth regressions exploit a linear relationship between expected returns and exposure to (priced) risk factors. 
The basic idea of the regression approach is to project asset returns on factor exposures or characteristics that might resemble exposure to a risk factor in the cross section. 
The estimates are then aggregated across time to test if a risk factor is priced. 
In this chapter, we present a simple implementation of @Fama1973 to introduce the concept of their regressions. We use 49 industry portfolios as test assets to estimate the risk premium associated with the three factors included in @Fama1993.

The Fama-MacBeth procedure is a simple two-step approach: 
The first step uses the exposures (characteristics) as explanatory variables in $T$ cross-sectional regressions, i.e.,
$$r_{i,t+1} = \alpha_i + \lambda^{M}_t \beta^M_{i,t}  + \lambda^{SMB}_t \beta^{SMB}_{i,t} + \lambda^{HML}_t \beta^{HML}_{i,t} + \epsilon_{i,t}\text{, for each t}.$$ 
Here, we are interested in the compensation $\lambda^{f}_t$ for the exposure to each risk factor $\beta^{f}_{i,t}$ at each time point, i.e., the risk premium. Note the terminology: $\beta^{f}_{i,t}$ is a asset-specific characteristic. *If* there is a linear relationship between expected returns and the characteristic in a given month, we expect the regression coefficient to reflect the relationship, i.e., $\lambda_t^{f}\neq0$. 
In a second step, the time-series average $\frac{1}{T}\sum\limits_{t=1}^T \hat\lambda^{f}_t$ of the estimates $\hat\lambda^{f}_t$ can then be interpreted as the risk premium for the specific risk factor $f$. We follow @Zaffaroni2022 and consider the standard cross-sectional regression to predict future returns. If the characteristics are replaced with time $t+1$ variables, the regression approach rather captures risk attributes. 

Before we move to the implementation, we want to highlight that the characteristics, e.g., $\hat\beta^{f}_{i}$, are typically estimated in a separate step. You might thus worry that the errors of $\hat\beta^{f}_{i}$ impact the risk premiums' standard errors. Measurement error in $\hat\beta^{f}_{i}$ indeed affects the risk premium estimates, i.e., they lead to biased estimates. The literature provides adjustments for this bias (see, e.g., @shanken1992, @kim1995, @chen2015, among others), but also shows that the bias goes to zero as $T \to \infty$. We refer to @gagliardini2016 for an indepth discussion covering also the case of time-varying betas and estimation on individual stocks. Moroever, if you plan to use Fama-MacBeth regressions with individual stocks: @hou2020 advocate using weighed-least squares to estimate the coefficients such that they are not biased towards small firms. Without this adjustment, the high number of small firms would drive the coefficient estimates.

## Data preparation

The current chapter relies on this set of packages. 
```{r}
library(tidyverse)
library(RSQLite)
library(lubridate)
library(broom)
library(sandwich)
```

We illustrate @Fama1973-regressions with the monthly CRSP sample and use three characteristics to explain the cross section of returns: market capitalization, the book-to-market ratio and the CAPM beta, the covariance of the excess stock returns with the market excess returns. We collect the data from our `SQLite`-database introduced in our chapter on *"Accessing & managing financial data"*.

```{r}
tidy_finance <- dbConnect(SQLite(), "data/tidy_finance.sqlite",
  extended_types = TRUE
)

crsp_monthly <- tbl(tidy_finance, "crsp_monthly") %>%
  collect()

compustat <- tbl(tidy_finance, "compustat") %>%
  collect()

beta <- tbl(tidy_finance, "beta") %>%
  collect()
```

We use the Compustat data to compute the book to market ratio and the (log) market capitalization. 
We also use the CAPM betas based on daily data which have been computed already in the previous Chapters.

```{r}
bm <- compustat %>%
  mutate(month = floor_date(ymd(datadate), "month")) %>%
  left_join(crsp_monthly, by = c("gvkey", "month")) %>%
  left_join(beta, by = c("permno", "month")) %>%
  transmute(
    gvkey,
    bm = be / mktcap,
    log_mktcap = log(mktcap),
    beta = beta_daily,
    sorting_date = month %m+% months(6)
  )

data_fama_macbeth <- crsp_monthly %>%
  left_join(bm, by = c("gvkey", "month" = "sorting_date")) %>%
  group_by(permno) %>%
  arrange(month) %>%
  fill(c(beta, bm, log_mktcap), .direction = "down") %>%
  ungroup() %>%
  left_join(crsp_monthly %>%
    select(permno, month, ret_excess_lead = ret) %>%
    mutate(month = month %m-% months(1)),
  by = c("permno", "month")
  ) %>%
  select(permno, month, ret_excess_lead, beta, log_mktcap, bm) %>%
  drop_na()
```

## Cross-sectional regression

Next, we run the cross-sectional regressions with the characteristics as explanatory variable for each month.  
We regress the returns of the test assets at a particular point in time on the characteristics of each asset. 
By doing so, we get an estimate of the risk premiums $\hat\lambda^{F_f}_t$ for each point in time. 

```{r}
risk_premiums <- data_fama_macbeth %>%
    nest(data = -month) %>% 
    mutate(estimates = map(.x = data, 
                        ~tidy(lm(ret_excess_lead ~ . - permno, data = .x)))) %>% 
  unnest(estimates)
```

## Time-series aggregation

Now that we have the risk premiums' estimates for each period, we can average across the time-series dimension to get the expected risk premium for each characteristic. Similarly, we manually create the t-test statistics for each regressor, which we can then compare to usual critical values of 1.96 or 2.576 for two-tailed significance tests. 

```{r}
price_of_risk <- risk_premiums %>%
  group_by(factor = term) %>%
  summarise(
    risk_premium = mean(estimate),
    t_statistic = risk_premium / sd(estimate) * sqrt(n())
  )
```

On a final note: It is common to adjust for autocorrelation in the time of prices of risk when reporting standard errors. This can be done by computing Newey West standard errors. One necessary input to compute Newey-West standard errors is a chosen bandwith based on the number of lags employed for the estimation. While it seems that often researchers default on choosing a pre-specified lag length of 6 months, we rather recommend a data driven approach. The appearent *default* approach for many Finance applications can be enforced in the code below by explicitly enforcing `NeweyWest(., lag = 6, prewhite = FALSE)` in the code below. 

```{r}
regressions_for_newey_west <- risk_premiums %>%
  select(month, factor = term, estimate) %>%
  nest(data = c(month, estimate)) %>%
  mutate(
    model = map(data, ~ lm(estimate ~ 1, .)),
    mean = map(model, tidy)
  )

price_of_risk_newey_west <- regressions_for_newey_west %>%
  mutate(newey_west_se = map_dbl(model, ~ sqrt(NeweyWest(.)))) %>%
  unnest(mean) %>%
  mutate(t_statistic_newey_west = estimate / newey_west_se) %>%
  select(factor,
    risk_premium = estimate,
    t_statistic_newey_west
  )

left_join(price_of_risk,
          price_of_risk_newey_west %>% select(factor, t_statistic_newey_west),
          by = "factor")
```

The results can be interpreted as follows: Higher book-to-market stocks earn higher expected future returns, the negative value for log market cap reflects the size premium for smaller stocks and the negative value for CAPM beta as a characteristic is in line with the well-established betting against beta anomalies, indicating that investors with borrowing constraints tilt their portfolio towards high beta stocks in order to replicate a levered market portfolio [@Frazzini2014].

## Exercises

1. Download a larger sample of test assets from Kenneth French's homepage and reevaluate the risk premiums for industry portfolios instead of individual stocks.
1. Implement a rolling-window regression for the time-series estimation of the factor exposure. Skip one month after each rolling period before including the exposures in the cross-sectional regression to avoid a look-ahead bias. Then, adapt the cross-sectional regression and compute the average risk premiums.