---
title: Working with Stock Returns
aliases:
  - ../introduction-to-tidy-finance.html
  - ../r/introduction-to-tidy-finance.html
  - ../python/introduction-to-tidy-finance.html
  - ../working-with-stock-returns.html
  - ../r/working-with-stock-returns.html
  - ../python/working-with-stock-returns.html
metadata:
  pagetitle: Working with Stock Returns
  description-meta: Learn how to use R and Python for downloading and analyzing stock market data.
---

```{python}
#| label: working-with-stock-returns-setup
#| echo: false
exec(open("assets/render-settings.py").read())
```

The main aim of this chapter is to familiarize yourself with the core packages for working with stock market data. We focus on downloading and visualizing stock data from data provider Yahoo Finance.\index{tidyverse}\index{polars}

At the start of each session, we load the required packages. Throughout the entire book, we use a core data frame library to handle our data. In this chapter, we also load the `tidyfinance` package to download stock price data. This package provides a convenient wrapper for various quantitative functions compatible with our core packages and our book.

We use the following packages throughout this chapter:

::: {.panel-tabset group="language"}
### R

Throughout the entire book, we always use the `tidyverse` [@Wickham2019] package. We also load the `tidyfinance` package to download stock price data and the `scales` [@scales] package, which provides nice formatting for axis labels in visualizations.

You typically have to install a package once before you can load it into your active R session. In case you have not done this yet, call, for instance, `install.packages("tidyfinance")`.\index{tidyfinance}

```{r}
#| label: working-with-stock-returns-packages
#| message: false
library(tidyverse)
library(tidyfinance)
library(scales)

theme_set(theme_minimal())
```

### Python

Throughout the entire book, we use the `polars` package [@polars] as our core data frame library. `polars` is a modern, high-performance alternative to `pandas` with an expressive, expression-based API and excellent handling of large datasets.\index{polars} We also load the `tidyfinance` package to download stock price data.

You typically have to install a package once before you can load it into your active Python session. In case you have not done this yet, call, for instance, `pip install polars tidyfinance` in your terminal.\index{tidyfinance}

```{python}
#| label: working-with-stock-returns-packages-py
import polars as pl
import tidyfinance as tf

tf.set_backend("polars")
```

Note that `import polars as pl` implies that we can call all polars functions later with a simple `pl.function()`. Instead, utilizing `from polars import *` is generally discouraged, as it leads to namespace pollution. This statement imports all functions and classes from `polars` into your current namespace, potentially causing conflicts with functions you define or those from other imported libraries. Using the `pl` abbreviation is a very convenient way to prevent this.\index{polars} With `tf.set_backend("polars")`, all `tidyfinance` download functions return `polars` data frames directly.\index{tidyfinance}

:::

## Downloading Data

We first download daily prices for one stock symbol, e.g., the Apple stock (*AAPL*), directly from the data provider Yahoo Finance. To download the data, you can use the function `download_data`.

In the following code, we request daily data from the beginning of 2000 to the end of the last year, which is a period of more than 20 years.\index{Stock prices}

::: {.panel-tabset group="language"}
### R

In case this is the first time you use the package `tidyfinance`, you may be asked once to install some additional packages in the process of downloading the data. If you do not know how to use it, make sure you read the help file by calling `?download_data`. We especially recommend taking a look at the examples section of the documentation.

```{r}
#| label: working-with-stock-returns-download-apple-prices
prices <- download_data(
  domain = "stock_prices",
  symbols = "AAPL",
  start_date = "2000-01-01",
  end_date = "2024-12-31"
)
prices
```

### Python

Because we set the `polars` backend above, `tf.download_data()` returns a `polars` data frame that we work with for the rest of the chapter. The `date` column arrives as the `polars` `Date` type — calendar dates without a time component — a convention we use throughout the book.

```{python}
#| label: working-with-stock-returns-download-apple-prices-py
prices = tf.download_data(
    domain="stock_prices",
    symbols="AAPL",
    start_date="2000-01-01",
    end_date="2024-12-31"
)
prices.head()
```

:::

\index{Data!Yahoo Finance} `download_data(domain = "stock_prices")` downloads stock market data from Yahoo Finance. The function returns a data frame with eight self-explanatory columns: `symbol`, `date`, the daily `volume` (in the number of traded shares), the market prices at the `open`, `low`, `high`, `close`, and the `adjusted_close` price in USD. The adjusted prices are corrected for anything that might affect the stock price after the market closes, e.g., stock splits and dividends. These actions affect the quoted prices, but they have no direct impact on the investors who hold the stock. Therefore, we often rely on adjusted prices when it comes to analyzing the returns an investor would have earned by holding the stock continuously.\index{Stock price adjustments}

We visualize the time series of adjusted prices in @fig-100 based on the principles of the grammar of graphics [@Wilkinson2012].\index{Graph!Time series}

::: {.panel-tabset group="language"}
### R

We use the `ggplot2` package [@ggplot2] to visualize the time series.

```{r}
#| label: fig-100
#| fig-cap: "Prices are in USD, adjusted for dividend payments and stock splits."
#| fig-alt: "Title: Apple stock prices between beginning of 2000 and end of 2024. The figure shows that the stock price of Apple increased from about 1 USD to around 250 USD."
prices |>
  ggplot(aes(x = date, y = adjusted_close)) +
  geom_line() +
  labs(
    x = NULL,
    y = NULL,
    title = "Apple stock prices between beginning of 2000 and end of 2024"
  )
```

### Python

We use the `plotnine` package [@plotnine] to visualize the time series. Note that generally, we do not recommend using the `*` import style. However, we use it here only for the plotting functions, which are distinct to `plotnine` and have very plotting-related names. So, the risk of misuse through a polluted namespace is marginal.

```{python}
#| label: working-with-stock-returns-import-plotnine
from plotnine import *

theme_set(theme_minimal())
```

Creating figures becomes very intuitive with the Grammar of Graphics, as the following code chunk demonstrates.

```{python}
#| label: wsr-fig-100-py
#| fig-cap: "Prices are in USD, adjusted for dividend payments and stock splits."
#| fig-alt: "Title: Apple stock prices between beginning of 2000 and end of 2024. The figure shows that the stock price of Apple increased from about 1 USD to around 250 USD."
apple_prices_figure = (
    ggplot(prices, aes(y="adjusted_close", x="date"))
    + geom_line()
    + labs(x="", y="", title="Apple stock prices between beginning of 2000 and end of 2024")
)
apple_prices_figure.show()
```

:::

## Computing Returns

Instead of analyzing prices, we compute daily returns defined as $r_t = p_t / p_{t-1} - 1$, where $p_t$ is the adjusted price at the end of day $t$.\index{Returns} In that context, computing the previous value (the lag) is helpful.

::: {.panel-tabset group="language"}
### R

The function `lag()` returns the previous value.

```{r}
#| label: working-with-stock-returns-compute-daily-returns
returns <- prices |>
  arrange(date) |>
  mutate(ret = adjusted_close / lag(adjusted_close) - 1) |>
  select(symbol, date, ret)
returns
```

### Python

```{python}
#| label: working-with-stock-returns-compute-daily-returns-py
returns = (prices
    .sort("date")
    .with_columns(ret=pl.col("adjusted_close").pct_change())
    .select("symbol", "date", "ret")
)
returns
```

:::

The resulting data frame has three columns, the last of which contains the daily returns (`ret`). Note that the first entry naturally contains a missing value because there is no previous price.\index{Missing value}

::: {.panel-tabset group="language"}
### R

Obviously, the use of `lag()` would be meaningless if the time series is not ordered by ascending dates.\index{Lag observations} The command `arrange()` provides a convenient way to order observations in the correct way for our application. If you want to order observations by descending values, you could, for instance, use `arrange(desc(ret))`. Always check that your data has the desired structure before calling `lag()` or similar functions.

### Python

Obviously, the use of `pct_change()` would be meaningless if the time series is not ordered by ascending dates. The method `sort()` provides a convenient way to order observations in the correct way for our application. In case you want to order observations by descending dates, you can use the parameter `descending=True`.

:::

For the upcoming examples, we remove missing values as these would require separate treatment for many applications. For example, missing values can affect sums and averages by reducing the number of valid data points if not properly accounted for. In general, always ensure you understand why missing values occur and carefully examine if you can simply get rid of these observations.

::: {.panel-tabset group="language"}
### R

```{r}
#| label: working-with-stock-returns-drop-missing-returns
returns <- returns |>
  drop_na(ret)
```

### Python

```{python}
#| label: working-with-stock-returns-drop-missing-returns-py
returns = returns.drop_nulls("ret")
```

:::

Next, we visualize the distribution of daily returns in a histogram in @fig-101. \index{Graph!Histogram} Additionally, we draw a dashed line that indicates the historical five percent quantile of the daily returns to the histogram, which is a crude proxy for the worst possible return of the stock with a probability of at most five percent. This quantile is closely connected to the (historical) value-at-risk, a risk measure commonly monitored by regulators.\index{Value-at-risk} We refer to @Tsay2010 for a more thorough introduction to the stylized facts of financial returns.

::: {.panel-tabset group="language"}
### R

```{r}
#| label: fig-101
#| fig-alt: "Title: Distribution of daily Apple stock returns in percent. The figure shows a histogram of daily returns. The range indicates a few large negative values, while the remaining returns are distributed around zero. The vertical line indicates that the historical five percent quantile of daily returns was around negative three percent."
#| fig-cap: "The dotted vertical line indicates the historical five percent quantile."
quantile_05 <- quantile(returns$ret, probs = 0.05)
returns |>
  ggplot(aes(x = ret)) +
  geom_histogram(bins = 100) +
  geom_vline(aes(xintercept = quantile_05),
    linetype = "dashed"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Distribution of daily Apple stock returns"
  ) +
  scale_x_continuous(labels = percent)
```

### Python

```{python}
#| label: wsr-fig-101-py
#| fig-alt: "Title: Distribution of daily Apple stock returns in percent. The figure shows a histogram of daily returns. The range indicates a few large negative values, while the remaining returns are distributed around zero. The vertical line indicates that the historical five percent quantile of daily returns was around negative three percent."
#| fig-cap: "The dotted vertical line indicates the historical five percent quantile."
from mizani.formatters import percent_format

quantile_05 = returns["ret"].quantile(0.05)

apple_returns_figure = (
    ggplot(returns, aes(x="ret"))
    + geom_histogram(bins=100)
    + geom_vline(aes(xintercept=quantile_05), linetype="dashed")
    + labs(x="", y="", title="Distribution of daily Apple stock returns")
    + scale_x_continuous(labels=percent_format())
)
apple_returns_figure.show()
```

:::

Here, `bins = 100` determines the number of bins used in the illustration and, hence, implicitly sets the width of the bins. Before proceeding, make sure you understand how to use the geom `geom_vline()` to add a dashed line that indicates the historical five percent quantile of the daily returns. Before proceeding with *any* data, a typical task is to compute and analyze the summary statistics for the main variables of interest.

::: {.panel-tabset group="language"}
### R

```{r}
#| label: working-with-stock-returns-summary-statistics
returns |>
  summarize(across(
    ret,
    list(
      daily_mean = mean,
      daily_sd = sd,
      daily_min = min,
      daily_max = max
    )
  ))
```

### Python

```{python}
#| label: working-with-stock-returns-summary-statistics-py
returns.select("ret").describe()
```

:::

We see that the maximum *daily* return was `{r} returns |> pull(ret) |> max() * 100` percent. Perhaps not surprisingly, the average daily return is close to but slightly above 0. In line with the illustration above, the large losses on the day with the minimum returns indicate a strong asymmetry in the distribution of returns.

You can also compute these summary statistics for each year individually by grouping the data by year. More specifically, the few lines of code below compute the summary statistics from above for individual groups of data defined by the values of the column year. The summary statistics, therefore, allow an eyeball analysis of the time-series dynamics of the daily return distribution.\index{Summary statistics}

::: {.panel-tabset group="language"}
### R

We impose `group_by(year = year(date))`, where the call `year(date)` returns the year.

```{r}
#| label: working-with-stock-returns-summary-statistics-by-year
returns |>
  group_by(year = year(date)) |>
  summarize(across(
    ret,
    list(
      daily_mean = mean,
      daily_sd = sd,
      daily_min = min,
      daily_max = max
    ),
    .names = "{.fn}"
  )) |>
  print(n = Inf)
```

In case you wonder, the additional argument `.names = "{.fn}"` in `across()` determines how to name the output columns. It acts as a placeholder that gets replaced by the name of the function being applied (e.g., mean, sd, min, max) when creating new column names. The specification is rather flexible and allows almost arbitrary column names, which can be useful for reporting. The `print()` function controls the R console's output options.

### Python

We group with `group_by(pl.col("date").dt.year())`, where the call `dt.year()` extracts the year. Unlike `pandas`, `polars` has no single `describe()` per group, so we list the desired statistics explicitly inside `agg()`.

```{python}
#| label: working-with-stock-returns-summary-statistics-by-year-py
(returns
    .group_by(pl.col("date").dt.year().alias("year"))
    .agg(
        count=pl.len(),
        mean=pl.col("ret").mean(),
        std=pl.col("ret").std(),
        min=pl.col("ret").min(),
        q25=pl.col("ret").quantile(0.25),
        median=pl.col("ret").median(),
        q75=pl.col("ret").quantile(0.75),
        max=pl.col("ret").max(),
    )
    .sort("year")
    .with_columns(pl.col(pl.Float64).round(3))
)
```

:::

## Scaling Up the Analysis

As a next step, we generalize the previous code so that all computations can handle an arbitrary number of symbols (e.g., all constituents of an index). Following tidy principles, it is quite easy to download the data, plot the price time series, and tabulate the summary statistics for an arbitrary number of assets.

This is where the `tidyverse` magic starts: Tidy data makes it extremely easy to generalize the computations from before to as many assets or groups as you like. The following code takes any number of symbols and automates the download as well as the plot of the price time series. In the end, we create the table of summary statistics for all assets at once. For this example, we analyze data from all current constituents of the [Dow Jones Industrial Average index.](https://en.wikipedia.org/wiki/Dow_Jones_Industrial_Average)\index{Data!Dow Jones Index}

We first download a table with Dow Jones constituents again using `download_data()`, but this time with `domain = "constituents"`.

::: {.panel-tabset group="language"}
### R

```{r}
#| label: working-with-stock-returns-download-dow-constituents
#| message: false
symbols <- download_data(
  domain = "constituents",
  index = "Dow Jones Industrial Average"
)
symbols
```

### Python

```{python}
#| label: working-with-stock-returns-download-dow-constituents-py
symbols = tf.download_data(
    domain="constituents",
    index="Dow Jones Industrial Average"
)
```

:::

Conveniently, `tidyfinance` provides the functionality to get all stock prices from an index for a specific point in time with a single call.\index{Exchange!NASDAQ}

::: {.panel-tabset group="language"}
### R

```{r}
#| label: working-with-stock-returns-download-dow-prices
#| output: false
prices_daily <- download_data(
  domain = "stock_prices",
  symbols = symbols$symbol,
  start_date = "2000-01-01",
  end_date = "2024-12-31"
)
```

### Python

```{python}
#| label: working-with-stock-returns-download-dow-prices-py
prices_daily = tf.download_data(
    domain="stock_prices",
    symbols=symbols["symbol"].to_list(),
    start_date="2000-01-01",
    end_date="2024-12-31"
)
```

:::

The resulting data frame contains `{r} nrow(prices_daily)` daily observations for `{r} unique(prices_daily$symbol)` different stocks. @fig-102 illustrates the time series of the downloaded *adjusted* prices for each of the constituents of the Dow index. Make sure you understand every single line of code! What are the arguments of `aes()`? Which alternative `geoms` could you use to visualize the time series? Hint: if you do not know the answers try to change the code to see what difference your intervention causes.

::: {.panel-tabset group="language"}
### R

```{r}
#| label: fig-102
#| fig-cap: "Prices in USD, adjusted for dividend payments and stock splits."
#| fig-alt: "Title: Stock prices of Dow Jones index constituents. The figure shows many time series with daily prices. The general trend seems positive for most stocks in the Dow Jones index."
prices_daily |>
  ggplot(aes(
    x = date,
    y = adjusted_close,
    color = symbol
  )) +
  geom_line() +
  labs(
    x = NULL,
    y = NULL,
    color = NULL,
    title = "Stock prices of Dow Jones index constituents"
  ) +
  theme(legend.position = "none")
```

### Python

```{python}
#| label: wsr-fig-102-py
#| fig-cap: "Prices in USD, adjusted for dividend payments and stock splits."
#| fig-alt: "Title: Stock prices of Dow Jones index constituents. The figure shows many time series with daily prices. The general trend seems positive for most stocks in the Dow Jones index."
prices_figure = (
    ggplot(prices_daily, aes(y="adjusted_close", x="date", color="symbol"))
    + geom_line()
    + scale_x_date(date_breaks="5 years", date_labels="%Y")
    + labs(x="", y="", color="", title="Stock prices of Dow Jones index constituents")
    + theme(legend_position="none")
)
prices_figure.show()
```

:::

Do you notice the small differences relative to the code we used before? All we needed to do to illustrate all stock symbols simultaneously is to include `color = symbol` in the `ggplot` aesthetics. In this way, we generate a separate line for each symbol. Of course, there are simply too many lines on this graph to identify the individual stocks properly, but it illustrates our point of how to generalize a specific analysis to an arbitrary number of subjects quite well.

The same holds for stock returns. The same logic also applies to the computation of summary statistics: grouping by `symbol` is the key to aggregating the time series into symbol-specific variables of interest.\index{Summary statistics}

::: {.panel-tabset group="language"}
### R

Before computing the returns, we use `group_by(symbol)` such that the `mutate()` command is performed for each symbol individually.

```{r}
#| label: working-with-stock-returns-multi-symbol-returns-summary
returns_daily <- prices_daily |>
  group_by(symbol) |>
  mutate(ret = adjusted_close / lag(adjusted_close) - 1) |>
  select(symbol, date, ret) |>
  drop_na(ret)

returns_daily |>
  group_by(symbol) |>
  summarize(across(
    ret,
    list(
      daily_mean = mean,
      daily_sd = sd,
      daily_min = min,
      daily_max = max
    ),
    .names = "{.fn}"
  )) |>
  print(n = Inf)
```

### Python

Before computing the returns, we sort by symbol and date and add `.over("symbol")` to the `pct_change()` expression, so that returns are computed within each symbol individually without bleeding across symbol boundaries.

```{python}
#| label: working-with-stock-returns-multi-symbol-returns-summary-py
returns_daily = (prices_daily
    .sort("symbol", "date")
    .with_columns(ret=pl.col("adjusted_close").pct_change().over("symbol"))
    .select("symbol", "date", "ret")
    .drop_nulls("ret")
)

(returns_daily
    .group_by("symbol")
    .agg(
        count=pl.len(),
        mean=pl.col("ret").mean(),
        std=pl.col("ret").std(),
        min=pl.col("ret").min(),
        q25=pl.col("ret").quantile(0.25),
        median=pl.col("ret").median(),
        q75=pl.col("ret").quantile(0.75),
        max=pl.col("ret").max(),
    )
    .sort("symbol")
    .with_columns(pl.col(pl.Float64).round(3))
)
```

:::

Note that you are now also equipped with all tools to download price data for *each* symbol listed in the S&P 500 index with the same number of lines of code. Just download the constituents with `domain = "constituents", index = "S&P 500"`, which provides you with a data frame that contains each symbol that is (currently) part of the S&P 500.\index{Data!SP 500} However, don't try this if you are not prepared to wait for a couple of minutes because this is quite some data to download!

## Different Frequencies

Financial data often exists at different frequencies due to varying reporting schedules, trading calendars, and economic data releases. For example, stock prices are typically recorded daily, while macroeconomic indicators such as GDP or inflation are reported monthly or quarterly. Additionally, some datasets are recorded only when transactions occur, resulting in irregular timestamps. To compare data meaningfully, we have to align different frequencies appropriately. For example, to compare returns across different frequencies, we use annualization techniques.

So far, we have worked with daily returns, but we can easily convert our data to other frequencies. Let's create monthly returns from our daily data:

::: {.panel-tabset group="language"}
### R

```{r}
#| label: working-with-stock-returns-compute-monthly-returns
returns_monthly <- returns_daily |>
  mutate(date = floor_date(date, "month")) |>
  group_by(symbol, date) |>
  summarize(
    ret = prod(1 + ret) - 1,
    .groups = "drop"
  )
```

### Python

```{python}
#| label: working-with-stock-returns-compute-monthly-returns-py
returns_monthly = (returns_daily
    .with_columns(date=pl.col("date").dt.truncate("1mo"))
    .group_by(["symbol", "date"])
    .agg(ret=(pl.col("ret") + 1).product() - 1)
    .sort("symbol", "date")
)
```

:::

In this code, we first group the data by symbol and month and then compute monthly returns by compounding the daily returns: $(1+r_1)(1+r_2)\ldots(1+r_n)-1$. To visualize how return characteristics change across different frequencies, we can compare histograms as in @fig-103:

::: {.panel-tabset group="language"}
### R

```{r}
#| label: fig-103
#| fig-cap: "Returns are based on prices adjusted for dividend payments and stock splits."
#| fig-alt: "Title: Distribution of Apple returns across different frequencies. The figure shows the distribution of daily and monthly returns in two separate facets."
apple_returns <- bind_rows(
  returns_daily |>
    filter(symbol == "AAPL") |>
    mutate(frequency = "Daily"),
  returns_monthly |>
    filter(symbol == "AAPL") |>
    mutate(frequency = "Monthly")
)

apple_returns |>
  ggplot(aes(x = ret, fill = frequency)) +
  geom_histogram(position = "identity", bins = 50) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Frequency",
    title = "Distribution of Apple returns across different frequencies"
  ) +
  scale_x_continuous(labels = percent) +
  theme_minimal() +
  facet_wrap(~ frequency, scales = "free") +
  theme(legend.position = "none")
```

### Python

```{python}
#| label: wsr-fig-103-py
#| fig-cap: "Returns are based on prices adjusted for dividend payments and stock splits."
#| fig-alt: "Title: Distribution of Apple returns across different frequencies. The figure shows the distribution of daily and monthly returns in two separate facets."
apple_daily = (returns_daily
    .filter(pl.col("symbol") == "AAPL")
    .with_columns(frequency=pl.lit("Daily"))
)

apple_monthly = (returns_monthly
    .filter(pl.col("symbol") == "AAPL")
    .with_columns(frequency=pl.lit("Monthly"))
)

apple_returns = pl.concat([apple_daily, apple_monthly])

apple_returns_figure = (
    ggplot(apple_returns, aes(x="ret", fill="frequency"))
    + geom_histogram(position="identity", bins=50)
    + labs(
        x="", y="", fill="Frequency",
        title="Distribution of Apple returns across different frequencies"
    )
    + scale_x_continuous(labels=percent_format())
    + facet_wrap("frequency", scales="free")
    + theme(legend_position="none")
)
apple_returns_figure.show()
```

:::

## Other Forms of Data Aggregation

Of course, aggregation across variables other than `symbol` or `date` can also make sense. For instance, suppose you are interested in answering questions like: Are days with high aggregate trading volume likely followed by days with high aggregate trading volume? To provide some initial analysis on this question, we take the downloaded data and compute aggregate daily trading volume for all Dow index constituents in USD. Recall that the column `volume` is denoted in the number of traded shares.\index{Trading volume} Thus, we multiply the trading volume with the daily adjusted closing price to get a proxy for the aggregate trading volume in USD. Scaling by `1e-9` denotes daily trading volume in billion USD.

::: {.panel-tabset group="language"}
### R

```{r}
#| label: fig-104
#| fig-cap: "Total daily trading volume in billion USD."
#| fig-alt: "Title: Aggregate daily trading volume of Dow Jones index constitutens. The figure shows a volatile time series of daily trading volume, ranging from about 8 in 2000 to around 90 in 2024, with a maximum of more than 200."
trading_volume <- prices_daily |>
  group_by(date) |>
  summarize(trading_volume = sum(volume * adjusted_close))

trading_volume |>
  ggplot(aes(x = date, y = trading_volume)) +
  geom_line() +
  labs(
    x = NULL, y = NULL,
    title = "Aggregate daily trading volume of Dow Jones index constituents"
  ) +
  scale_y_continuous(labels = unit_format(unit = "B", scale = 1e-9))
```

### Python

```{python}
#| label: wsr-fig-104-py
#| fig-cap: "Total daily trading volume in billion USD."
#| fig-alt: "Title: Aggregate daily trading volume of Dow Jones index constitutens. The figure shows a volatile time series of daily trading volume, ranging from about 8 in 2000 to around 90 in 2024, with a maximum of more than 200."
trading_volume = (prices_daily
    .with_columns(trading_volume=(pl.col("volume") * pl.col("adjusted_close")) / 1e9)
    .group_by("date")
    .agg(pl.col("trading_volume").sum())
    .sort("date")
    .with_columns(trading_volume_lag=pl.col("trading_volume").shift(1))
)

trading_volume_figure = (
    ggplot(trading_volume, aes(x="date", y="trading_volume"))
    + geom_line()
    + scale_x_date(date_breaks="5 years", date_labels="%Y")
    + labs(
        x="", y="",
        title="Aggregate daily trading volume of Dow Jones index constituents"
        )
)
trading_volume_figure.show()
```

:::

@fig-104 indicates a clear upward trend in aggregated daily trading volume. In particular, since the outbreak of the COVID-19 pandemic, markets have processed substantial trading volumes, as analyzed, for instance, by @Goldstein2021.\index{Covid 19} One way to illustrate the persistence of trading volume would be to plot volume on day $t$ against volume on day $t-1$ as in the example below. In @fig-105, we add a dotted 45°-line to indicate a hypothetical one-to-one relation by `geom_abline()`, addressing potential differences in the axes' scales.

::: {.panel-tabset group="language"}
### R

```{r}
#| label: fig-105
#| fig-cap: "Total daily trading volume in billion USD."
#| fig-alt: "Title: Persistence in daily trading volume of Dow Jones index constituents. The figure shows a scatterplot where aggregate trading volume and previous-day aggregate trading volume neatly line up along a 45-degree line."
trading_volume |>
  ggplot(aes(x = lag(trading_volume), y = trading_volume)) +
  geom_point() +
  geom_abline(aes(intercept = 0, slope = 1),
    linetype = "dashed"
  ) +
  labs(
    x = "Previous day aggregate trading volume",
    y = "Aggregate trading volume",
    title = "Persistence in daily trading volume of Dow Jones index constituents"
  ) +
  scale_x_continuous(labels = unit_format(unit = "B", scale = 1e-9)) +
  scale_y_continuous(labels = unit_format(unit = "B", scale = 1e-9))
```

Do you understand where the warning `## Warning: Removed 1 rows containing missing values (geom_point).` comes from and what it means?\index{Error message}

### Python

```{python}
#| label: wsr-fig-105-py
#| fig-cap: "Total daily trading volume in billion USD."
#| fig-alt: "Title: Persistence in daily trading volume of Dow Jones index constituents. The figure shows a scatterplot where aggregate trading volume and previous-day aggregate trading volume neatly line up along a 45-degree line."
persistence_figure = (
    ggplot(trading_volume, aes(x="trading_volume_lag", y="trading_volume"))
    + geom_point()
    + geom_abline(aes(intercept=0, slope=1), linetype="dashed")
    + labs(
        x="Previous day aggregate trading volume",
        y="Aggregate trading volume",
        title="Persistence in daily trading volume of Dow Jones index constituents"
        )
)
persistence_figure.show()
```

:::

Purely eye-balling reveals that days with high trading volume are often followed by similarly high trading volume days.

## Key Takeaways

- You can use the `tidyverse` (R) or `polars` (Python) to efficiently analyze stock market data using consistent and scalable workflows.
- The `tidyfinance` package allows seamless downloading of historical stock prices and index data directly from Yahoo Finance.
- Tidy data principles enable efficient analysis of financial data.
- Adjusted stock prices provide a more accurate reflection of investor returns by accounting for dividends and stock splits.
- Summary statistics such as mean, standard deviation, and quantiles offer insights into stock return behavior over time.
- Visualizations created with `ggplot2` (R) or `plotnine` (Python) help identify trends, volatility, and return distributions in financial time series.
- Tidy data principles make it easy to scale financial analyses from a single stock to entire indices like the Dow Jones or S&P 500.
- Consistent workflows form the foundation for advanced financial analysis.

## Exercises

1.  Download daily prices for another stock market symbol of your choice from Yahoo Finance using `download_data()` from the `tidyfinance` package. Plot two time series of the symbol’s un-adjusted and adjusted closing prices. Explain any visible differences.
2.  Compute daily net returns for an asset of your choice and visualize the distribution of daily returns in a histogram using 100 bins. Also, use `geom_vline()` to add a dashed red vertical line that indicates the 5 percent quantile of the daily returns. Compute summary statistics (mean, standard deviation, minimum, and maximum) for the daily returns.
3.  Take your code from the previous exercises and generalize it such that you can perform all the computations for an arbitrary number of symbols (e.g., `symbol <- c("AAPL", "MMM", "BA")`). Automate the download, the plot of the price time series, and create a table of return summary statistics for this arbitrary number of assets.
4.  To facilitate the computation of the annualization factor, write a function that takes a vector of return dates as input and determines the frequency before returning the appropriate annualization factor.
5.  Are days with high aggregate trading volume often also days with large absolute returns? Find an appropriate visualization to analyze the question using the symbol `AAPL`.
