---
title: Discounted Cash Flow Analysis
metadata:
  pagetitle: DCF with R
  description-meta: Learn how to use the programming language R to value companies using discounted cash flow analysis.
---

In this chapter, we address a fundamental question: What is the value of a company? Company valuation is a critical tool that helps us determine the economic value of a business. Understanding a company's value is essential, whether it is for investment decisions, mergers and acquisitions, or financial reporting. Valuation is not just about assigning a number, as it is also a framework for making informed decisions. For example, investors use valuation to identify whether a stock is under- or overvalued, and companies rely on valuation for strategic decisions, like pricing an acquisition or preparing for an IPO.

Company valuation methods broadly fall into three categories:\index{Valuation methods}

- Market-based approaches compare companies using relative metrics like Price-to-Earnings ratios.
- Asset-based methods focus on the net value of a company's tangible and intangible assets.
- Income-based techniques value companies based on their ability to generate future cash flows.

We focus on Discounted Cash Flow (DCF) analysis, an income-based approach, because it captures three crucial aspects of valuation:\index{Discounted Cash Flow}\index{DCF} First, DCF explicitly accounts for the time value of money - the principle that a dollar today is worth more than a dollar in the future. By discounting future cash flows to present value with the appropriate discount rate, we incorporate both time preferences and risk.\index{Time value of money} Second, DCF is inherently forward-looking, making it particularly suitable for companies where historical performance may not reflect future potential. This characteristic is especially relevant when valuing growth companies or analyzing new business opportunities. Third, DCF analysis is flexible enough to accommodate various business models and capital structures, making it applicable across different industries and company sizes.

In our exposition of the DCF valuation framework, we focus on its three key components:\index{DCF structure}

- Free Cash Flow (FCF) forecasts represent the expected future cash available for distribution to investors after accounting for operating expenses, taxes, and investments.\index{Free Cash Flow}
- Terminal value captures the company's value beyond the explicit forecast period, often representing a significant portion of the total valuation.\index{Terminal value}
- Discount rate, typically the Weighted Average Cost of Capital (WACC), adjusts future cash flows to present value by incorporating risk and capital structure considerations.\index{WACC}

We make a few simplifying assumptions due to the diverse nature of businesses, which a simple model cannot cover. After all, there are entire textbooks written just on valuation. In particular, we assume that firms only conduct operating activities (i.e., financial statements do not include non-operating items), implying that firms do not own any assets that do not produce operating cash flows. Otherwise, you must value these non-operating activities separately for many real-world examples. 

In this chapter, we rely on the following packages to build a simple DCF analysis:

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(tidyfinance)
library(scales)
library(fmpapi)
```

::: {.callout-tip}
The `fmpapi` package is developed by Christoph Scheuch and not sponsored by or affiliated with FMP. However, you can get 15% off your FMP subscription by using this [affiliate link](https://site.financialmodelingprep.com/pricing-plans?couponCode=tidyfinance). By signing up through this link, you also support the development of this package at no extra cost to you.
:::

## Prepare Data

Before we can perform a DCF analysis, we need historical financial data to inform our forecasts. We use the Financial Modeling Prep (FMP) API to download financial statements. The `fmpapi` package provides a convenient interface for accessing this data in a tidy format.\index{Data!FMP API}

```{r}
#| cache: true
symbol <- "MSFT"

income_statements <- fmp_get(
  "income-statement", symbol, list(period = "annual", limit = 5)
)
cash_flow_statements <- fmp_get(
  "cash-flow-statement", symbol, list(period = "annual", limit = 5)
)
```

Our analysis centers on Free Cash Flow (FCF), which represents the cash available to all investors after the company has covered its operational needs and capital investments.\index{Free Cash Flow} We calculate FCF using the following formula:

$$\text{FCF} = \text{EBIT} + \text{Depreciation & Amortization} - \text{Taxes} + \Delta \text{Working Capital} - \text{CAPEX}$$
Each component of this formula serves a specific purpose in capturing the company's cash-generating ability:\index{Free Cash Flow components}

- EBIT (Earnings Before Interest and Taxes) measures core operating profit
- Depreciation & Amortization accounts for non-cash expenses
- Taxes reflect actual cash payments to tax authorities
- Changes in Working Capital capture cash tied up in operations
- Capital Expenditures (CAPEX) represent investments in long-term assets

We can implement this calculation by combining and transforming our financial statement data:

```{r}
dcf_data <- income_statements |> 
  mutate(
    ebit = net_income + income_tax_expense - interest_expense - interest_income
  ) |> 
  select(
    year = calendar_year, 
    ebit, revenue, depreciation_and_amortization, taxes = income_tax_expense
  ) |> 
  left_join(
    cash_flow_statements |> 
      select(
        year = calendar_year, 
        delta_working_capital = change_in_working_capital,
        capex = capital_expenditure
      ), join_by(year)
  ) |> 
  mutate(
    fcf = ebit + depreciation_and_amortization - taxes + delta_working_capital - capex
  ) |> 
  arrange(year) 
```

## Forecast Free-Cash-Flow

After calculating historical FCF, we need to project it into the future. While historical data provides a foundation, forecasting requires both quantitative analysis and qualitative judgment. We use a ratio-based approach that links all FCF components to revenue growth, making our forecasts more tractable.\index{Free Cash Flow!Forecasting}

First, we express each FCF component as a ratio relative to revenue. This standardization helps identify trends and makes forecasting more systematic. @fig-500 shows the historical evolution of these key financial ratios.

```{r}
#| label: fig-500
#| fig-cap: "Ratios are based on financial statements as provided through the FMP API." 
#| fig-alt: "Title: Key financial ratios of Microsoft between 2020 and 2024. The figure shows a line chart with years on the horizontal axis and financial ratios on the vertical axis."
dcf_data <- dcf_data |> 
  mutate(
    revenue_growth = revenue / lag(revenue) - 1,
    operating_margin = ebit / revenue,
    da_margin = depreciation_and_amortization / revenue,
    taxes_to_revenue = taxes / revenue,
    delta_working_capital_to_revenue =  delta_working_capital / revenue,
    capex_to_revenue = capex / revenue
  )

fig_financial_ratios <- dcf_data |> 
  pivot_longer(cols = c(operating_margin:capex_to_revenue)) |>
  ggplot(aes(x = year, y = value, color = name)) +
  geom_line() +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = percent) +
  labs(
    x = NULL, y = NULL, color = NULL,
    title = "Key financial ratios of Microsoft between 2020 and 2024"
  )
fig_financial_ratios
```

The operating margin, for instance, represents how much of each revenue dollar translates into operating profit (EBIT), while the CAPEX-to-revenue ratio indicates the company's investment intensity.\index{Financial ratios}

For our DCF analysis, we need to project these ratios into the future. These projections should reflect both historical patterns and forward-looking considerations such as: Industry trends and competitive dynamics, company-specific growth initiatives, expected operational efficiency improvements, planned capital investments, or working capital management strategies.

We demonstrate this forecasting approach in @fig-501. Note that our forecasted ratio dynamics just serve as an example and are not grounded in any economic analysis.

```{r}
#| label: fig-501
#| fig-cap: "Realized ratios are based on financial statements as provided through the FMP API, while forecasts are manually defined." 
#| fig-alt: "Title: Key financial ratios and ad-hoc forecasts of Microsoft between 2020 and 2029. The figure shows a line chart with years on the horizontal axis and financial ratios and their forecasts on the vertical axis."
dcf_data_forecast_ratios <- tribble(
  ~year, ~operating_margin, ~da_margin, ~taxes_to_revenue, ~delta_working_capital_to_revenue, ~capex_to_revenue,
  2025, 0.41, 0.09, 0.08, 0.001, -0.2,
  2026, 0.42, 0.09, 0.07, 0.001, -0.22,
  2027, 0.43, 0.09, 0.06, 0.001, -0.2,
  2028, 0.44, 0.09, 0.06, 0.001, -0.18,
  2029, 0.45, 0.09, 0.06, 0.001, -0.16
) |> 
  mutate(type = "Forecast")

dcf_data <- dcf_data |> 
  mutate(type = "Realized") |> 
  bind_rows(dcf_data_forecast_ratios)

fig_financial_ratios_forecast <- dcf_data |> 
  pivot_longer(cols = c(operating_margin:capex_to_revenue)) |> 
  ggplot(aes(x = year, y = value, color = name, linetype = rev(type))) +
  geom_line() +
    scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = percent) +
  labs(
    x = NULL, y = NULL, color = NULL, linetype = NULL,
    title = "Key financial ratios and ad-hoc forecasts of Microsoft between 2020 and 2029"
  )
fig_financial_ratios_forecast
```

The final step in our FCF forecast is projecting revenue growth. While there are multiple approaches to this task, we demonstrate a GDP-based method that links company growth to macroeconomic forecasts.\index{Revenue growth}

We use GDP growth forecasts from the [IMF World Economic Outlook (WEO)](https://www.imf.org/en/Publications/WEO/weo-database/2024/October/weo-report?c=111,&s=NGDP_RPCH,&sy=2020&ey=2029&ssm=0&scsm=1&scc=0&ssd=1&ssc=0&sic=1&sort=country&ds=.&br=1) database as our baseline. The WEO provides comprehensive economic projections, though it's important to note that these forecasts are periodically revised as new data becomes available.\index{Data!IMF WEO}

```{r}
gdp_growth <- tibble(
  year = 2020:2029,
  gdp_growth = c(-0.02163, 0.06055,	0.02512, 0.02887, 0.02765, 0.02153, 0.02028, 0.02120, 0.02122, 0.02122)
)

dcf_data <- dcf_data |> 
  left_join(gdp_growth, join_by(year))
```

Our approach models revenue growth as a linear function of GDP growth. This relationship captures the intuition that company revenues often move in tandem with broader economic activity, though usually with different sensitivity.\index{Growth modeling}

```{r}
revenue_growth_model <- dcf_data |> 
  lm(revenue_growth ~ gdp_growth, data = _) |> 
  coefficients()
 
dcf_data <- dcf_data |> 
  mutate(
    revenue_growth_modeled = revenue_growth_model[1] + revenue_growth_model[2] * gdp_growth,
    revenue_growth = if_else(type == "Forecast", revenue_growth_modeled, revenue_growth)  
  ) 
```

The model estimates two parameters: (i) an intercept that captures the company's baseline growth, and (ii) a slope coefficient that measures the company's sensitivity to GDP changes.

We visualize the historical and projected growth rates using this approach in @fig-502:

```{r}
#| label: fig-502
#| fig-cap: "Realized revue growth rates are based on financial statements as provided through the FMP API, while forecasts are modeled unsing IMF WEO forecasts." 
#| fig-alt: "Title: GDP growth and Microsoft revenue growth and modeled forecasts between 2020 and 2029. The figure shows a line chart with years on the horizontal axis and gdp growth, revenue growth and their forecasts on the vertical axis."
fig_growth <- dcf_data |> 
  filter(year >= 2021) |> 
  pivot_longer(cols = c(revenue_growth, gdp_growth)) |> 
  ggplot(aes(x = year, y = value, color = name, linetype = rev(type))) +
  geom_line() +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = percent) +
  labs(
    x = NULL, y = NULL, color = NULL, linetype = NULL,
    title = "GDP growth and Microsoft revenue growth and modeled forecasts between 2020 and 2029"
  )
fig_growth
```

While more sophisticated approaches exist, including proprietary analyst forecasts or bottom-up market analysis, this method provides a transparent and data-driven starting point for revenue projections.\index{Forecasting methods}

With all components in place - revenue growth projections and financial ratios - we can now calculate our FCF forecasts. We first need to convert our growth rates into revenue projections and then apply our forecasted ratios to compute each FCF component.\index{Free Cash Flow!Computation}

```{r}
dcf_data$revenue_growth[1] <- 0
dcf_data$revenue <- dcf_data$revenue[1] * cumprod(1 + dcf_data$revenue_growth)

dcf_data <- dcf_data |> 
  mutate(
    ebit = operating_margin * revenue,
    depreciation_and_amortization = da_margin * revenue,
    taxes = taxes_to_revenue * revenue,
    delta_working_capital = delta_working_capital_to_revenue * revenue,
    capex = capex_to_revenue * revenue,
    fcf = ebit + depreciation_and_amortization - taxes + delta_working_capital - capex
  )
```

We visualize the resulting FCF projections in @fig-503:

```{r}
#| label: fig-503
#| fig-cap: "Realized growth rates are based on financial statements as provided through the FMP API, while forecasts are manually defined." 
#| fig-alt: "Title: GDP growth and Microsoft revenue growth and modeled forecasts between 2020 and 2029. The figure shows a line chart with years on the horizontal axis and gdp growth, revenue growth and their forecasts on the vertical axis."
fig_fcf <- dcf_data |>
  ggplot(aes(x = year, y = fcf / 1e9)) +
  geom_col(aes(fill = type)) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma) + 
  labs(
    x = NULL, y = "Free Cash Flow (in B USD)", fill = NULL,
    title = "Actual and predicted free cash flow for Microsoft from 2020 to 2029"
  )
fig_fcf
```

## Continuation Value

A key component of DCF analysis is the continuation value (or terminal value), which represents the company's value beyond the explicit forecast period. This value often constitutes the majority of the total valuation, making its estimation particularly important.\index{Continuation value}
The most common approach is the Perpetuity Growth Model, which assumes FCF grows at a constant rate indefinitely. The formula for this model is:

$$TV_{T} = \frac{FCF_{T+1}}{r - g}$$

where $TV_{T}$ is the terminal value at time $T$, $FCF_{T+1}$ is the free cash flow in the first year after our forecast period, $r$ is the discount rate (typically WACC, see below), and $g$ is the perpetual growth rate.

The perpetual growth rate $g$ should reflect long-term economic growth potential. A common benchmark is the long-term GDP growth rate, as few companies can sustainably grow faster than the overall economy indefinitely.\index{Growth rate!Perpetual} For instance, last 20 years GDP growth is a sensible assumption (nominal growth rate is 4% for the US).

Let's implement the perpetuity growth model:

```{r}
compute_terminal_value <- function(last_fcf, growth_rate, discount_rate){
  last_fcf * (1 + growth_rate) / (discount_rate - growth_rate)
}

last_fcf <- tail(dcf_data$fcf, 1)
terminal_value <- compute_terminal_value(last_fcf, 0.04, 0.08)
terminal_value / 1e9
```

Note that while we use the Perpetuity Growth Model here, practitioners often cross-check their estimates with alternative methods like the exit multiple approach, which bases the terminal value on comparable company valuations.^[See (corporatefinanceinstitute.com/)[https://corporatefinanceinstitute.com/resources/valuation/exit-multiple/)] for an intuitive explanation of the exit multiple approach.]
\index{Valuation methods!Exit multiple}

## Discount Rates

The final component of our DCF analysis involves discounting future cash flows to present value. We typically use the Weighted Average Cost of Capital (WACC) as the discount rate, as it represents the blended cost of financing for all company stakeholders.\index{WACC}

The WACC formula combines the costs of equity and debt financing:

$$WACC = \frac{E}{D+E} \cdot r^E + \frac{D}{D+E} \cdot r^D \cdot (1 - \tau),$$

where $E$ is the market value of the company’s equity with required return $r^E$, $D$ is the market value of the company's debt with pre-tax return $r^D$, and $\tau$ is the tax rate.

While you can often find estimates of WACC from financial databases or analysts’ reports, sometimes you may need to calculate it yourself. Let’s walk through the practical steps to estimate WACC using real-world data: 
- $E$ is typically measured as the market value of the company’s equity. One common approach is to calculate it by subtracting net debt (total debt minus cash) from the enterprise value.
- $D$ is often measured using the book value of the company’s debt. While this might not perfectly reflect market conditions, it’s a practical starting point when market data is unavailable.
- The Capital Asset Pricing Model (CAPM) is a popular method to estimate the cost of equity $r^E$. It considers the risk-free rate, the equity risk premium, and the company’s beta. For a detailed guide on how to estimate the CAPM, we refer to Chapter [Capital Asset Pricing Model](capital-asset-pricing-model.qmd).
- The return on debt $r^D$ can also be estimated in different ways. For instance, effective interest rates can be calculated as the ratio of interest expense to total debt from financial statements. This gives you a real-world measure of what the company is currently paying. Alternatively, you can look up corporate bond spreads for companies in the same rating group. For highly rated companies like Microsoft, this would reflect their low-risk profile and correspondingly low borrowing costs.

If you'd rather not estimate WACC manually, there are excellent resources available to help you find industry-specific discount rates. One of the most widely used sources is Aswath Damodaran’s [database](https://pages.stern.nyu.edu/~adamodar/New_Home_Page/datacurrent.html). He maintains an extensive database that provides a wealth of financial data, including estimated discount rates, cash flows, growth rates, multiples, and more. What makes his database particularly valuable is its level of detail and coverage of multiple industries and regions. For example, if you’re analyzing a company in the Computer Services sector, as we do here, you can look up the industry’s average WACC and use it as a benchmark for your analysis. The following code chunk downloads the WACC data and extracts the value for this industry:

```{r}
#| cache: true
library(readxl)

file <- tempfile(fileext = "xls")

url <- "https://pages.stern.nyu.edu/~adamodar/pc/datasets/wacc.xls"
download.file(url, file)
wacc_raw <- read_xls(file, sheet = 2, skip = 18)
unlink(file)

wacc <- wacc_raw |> 
  filter(`Industry Name` == "Computer Services") |> 
  pull(`Cost of Capital`)
```

## Compute DCF Value

Having established all components, we can now compute the total company value. The DCF value combines two elements:\index{DCF!Computation}

- The present value of explicit forecast period cash flows
- The present value of the terminal value

This is expressed mathematically as:

$$
\text{Total DCF Value} = \sum_{t=1}^{T} \frac{\text{FCF}t}{(1 + \text{WACC})^t} + \frac{\text{TV}{T}}{(1 + \text{WACC})^T}
$$

where $T$ is the length of our forecast period. Let's implement this calculation:

```{r}
forecasted_years <- 5

compute_dcf <- function(wacc, growth_rate, years = 5) {
  free_cash_flow <- dcf_data$fcf
  last_fcf <- tail(free_cash_flow, 1)
  terminal_value <- compute_terminal_value(last_fcf, growth_rate, wacc)
  
  present_value_fcf <- free_cash_flow / (1 + wacc)^(1:years)
  present_value_tv <- terminal_value / (1 + wacc)^years
  total_dcf_value <- sum(present_value_fcf) + present_value_tv
  total_dcf_value
}

compute_dcf(wacc, 0.03) / 1e9
```

Note that this valuation represents an enterprise value - the total value of the company's operations. To arrive at an equity value, we would need to subtract net debt (total debt minus cash and equivalents).\index{Enterprise value}

## Sensitvity Analysis

DCF valuation is only as robust as its underlying assumptions. Given the inherent uncertainty in forecasting, it's crucial to understand how changes in key inputs affect our valuation.\index{Sensitivity analysis}

While we could examine sensitivity to various inputs like operating margins or capital expenditure ratios, we focus on two critical drivers:

- The perpetual growth rate, which determines long-term value creation
- The WACC, which affects how we value future cash flows

Let's implement a sensitivity analysis that varies these parameters:

```{r}
#| label: fig-504
#| fig-cap: "DCF value combines data from FMP API, ad-hoc forecasts of financial ratios, and IMF WEO growth forecasts." 
#| fig-alt: "Title: DCF value of Microsoft for different WACC and growth scenarios. The figure shows a tile chart different values of WACC on the horizontal axis and perpetual growth rates on the vertical axis. Each tile shows a corresponding DCF value, illustrating the sensitivity of the DCF analysis to assumptions."
wacc_range <- seq(0.06, 0.08, by = 0.01)
growth_rate_range <- seq(0.02, 0.04, by = 0.01)

sensitivity <- expand_grid(
  wacc = wacc_range,
  growth_rate = growth_rate_range
) |>
  mutate(value = pmap_dbl(list(wacc, growth_rate), compute_dcf))

fig_sensitivity <- sensitivity |> 
  mutate(value = round(value / 1e9, 0)) |> 
  ggplot(aes(x = wacc, y = growth_rate, fill = value)) +
  geom_tile() +
  geom_text(aes(label = comma(value)), color = "white") +
  scale_x_continuous(labels = percent) + 
  scale_y_continuous(labels = percent) +
  scale_fill_continuous(labels = comma) + 
  labs(
    title = "DCF value of Microsoft for different WACC and growth scenarios",
    x = "WACC",
    y = "Perpetual growth rate",
    fill = "Company value"
  ) + 
  guides(fill = guide_colorbar(barwidth = 15, barheight = 0.5))
fig_sensitivity
```

@fig-504 reveals several key insights about our valuation:\index{Visualization!Sensitivity} The valuation is highly sensitive to both WACC and growth assumptions; small changes in either parameter can lead to substantial changes in company value; the relationship between these parameters and company value is non-linear; and the impact of growth rate changes becomes more pronounced at lower WACCs.

## From DCF to Equity Value

Our DCF analysis yields the enterprise value - the value of the company's operations. To arrive at the equity value that belongs to shareholders, we need to make several adjustments:\index{Equity value}

$$\text{Equity Value} = \text{DCF Value} + \text{Non-Operating Assets} - \text{Value of Debt}$$
where non-operating assets are not essential to operations, but generate income (e.g., marketable securities, vacant land, idle equipment) and the value of debt is in theory the market value of total debt, but in practice typically book debt. We leave it as an exercise to calculate the equity value using the DCF value from above. 

## Key takeaways

Discounted Cash Flow analysis provides a structured approach to company valuation. In this chapter, you learned to:

- Calculate Free Cash Flow using financial statement data and ratio analysis.
- Project future cash flows by combining company-specific metrics with macroeconomic forecasts.
- Estimate terminal value using the perpetuity growth method.
- Apply appropriate discount rates based on industry benchmarks.
- Test the sensitivity of valuations to key assumptions.
- Convert enterprise values to equity values using balance sheet adjustments.

## Exercises

1. Download financial statements for another company of your choice and compute its historical Free Cash Flow. Compare the results with the Microsoft example from this chapter.
1. Create a function that automatically generates FCF forecasts using different sets of ratio assumptions. Use it to create alternative scenarios for Microsoft.
1. Implement an exit multiple approach for terminal value calculation and compare the results with the perpetuity growth method.
1. Extend the sensitivity analysis to include operating margin assumptions. Create a visualization showing how changes in margins affect the final valuation.
1. Calculate the equity value of Microsoft by adjusting the CDF value as described above. 
